---
title: "Joined_Data"
author: "Veronica Strempfer"
date: "2026-03-13"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Packages 
```{r}
# joined csv files from shared drive with FIPs
library(dplyr)
library(tidyverse)
library(readr)
library(readxl)
```

#Census Tract Data and Unique IDS
```{r, include=FALSE}
#this is me reading in the Census tract data
census_tracts <- read.csv("CA AMI Census Tracts 2022.csv")

#Los Angeles County Data Extracted from the larger dataset
LA <- census_tracts %>% 
  filter(substr(FIP, 1, 4) == "6037") 

#Identifying the unique LA Tract IDS there are 2467
UniLAC <-  LA %>%
  filter(str_starts(as.character(FIP), "6037")) %>%
  distinct(FIP)

```

# City Data and Tract Overlap Data

```{r, include=FALSE}
#this is me reading in the city data with city ID and name 
cities <- read_excel("LEAD Tool Cities List_TFIPS_Overlap.xlsx", sheet="CA Cities")
#this is me removing StateFIPS from City Data set
cities<- cities %>% 
  select(-1)
#this is me Tract FIPS, City ID
tracts <- read_excel("LEAD Tool Cities List_TFIPS_Overlap.xlsx", sheet= "TOverlap")

#this is me filtering out only CA Cities 
CATractsdf <- tracts %>% 
  filter(substr(`State FIPS`, 1,1) == "6") 

#this is me filtering out only CA Cities there are 2352
CACityTractdf <- CATractsdf %>% 
  filter(substr(`Tract FIPS`, 1, 4) == "6037") 


```

#Joining Data 

```{r, include=FALSE}

#this is me joining the CACityTractdf to cities data
CACityTractdf <- CACityTractdf %>%
  left_join(cities %>% select('City ID', Name), by = c("City ID" = "City ID")) %>%
  rename(City_Name = Name)

#this is me searching for cities with strange/special characters 
cities_special <- CACityTractdf %>%
  filter(if_any( City_Name,
                ~ str_detect(as.character(.), "[^[:alnum:] ]")))
#this is me changing La Canada Flintridge
CACityTractdf <- CACityTractdf%>%  
 mutate(City_Name = str_replace(
    City_Name,
    "La Cañada Flintridge",
    "La Canada Flintridge"
  ))

#this is me removing State FIPS from the City Tract df
CACityTractdf <- CACityTractdf%>%  
  select(-c(1,3,4))

#this is me renaming Tract FIPS to FIP
CACityTractdf <- CACityTractdf %>%  
   rename(FIP = `Tract FIPS`)
#this is me merging data based on FIP. Every FIP will be associated with a City Name 
LACityTractdf <- LA %>%
  left_join(CACityTractdf, by = "FIP") 

#this is me saving the data set LACityData 
write.csv(LACityTractdf, "LACityData.csv", row.names = FALSE)

CACityTractdf[, ]
```

#Loading the new Dataset 
```{r, include=FALSE}
# loading the new merged data to find factor information 
LACityData <- read_csv("LACityData.csv")

#this is me making City Data into a Factor 
LACityData$City_Name<- as.factor(LACityData$City_Name)

#this is me getting the frequency of City Names 
table(LACityData$City_Name)
```



```{r}
#this is me getting the frequency of City Names 
table(LACityData$City_Name)
```

```{r, include=FALSE}
library(readr)
library(dplyr)
tree_raw <- read_csv("Public_Works_Road_Maintenance_Division_Tree_Inventory_(Public_View)_2479203802696342889.csv")

unique(tree_raw$COMMUNITY)
```

```{r, include=FALSE}
library(dplyr)
library(stringr)
# filter by city (ie Arcadia) create into data frame and sum over that way
# cities from joined data included here: Arcadia, Azusa, Cerritos, Claremont, Covina, Duarte, Glendora, Hawthorne, La Habra Heights, La Verne, Lakewood, Lancaster, Long Beach, Los Angeles, Monrovia, Palmdale, Pomona, San Dimas, Santa Clarita, Sierra Madre, South El Monte, Walnut, Whittier

tree_clean <- tree_raw %>%
  filter(!is.na(COMMUNITY)) %>%
  mutate(
    COMMUNITY = toupper(str_squish(COMMUNITY))
  )
tree_clean %>%
  count(COMMUNITY, sort = TRUE)
tree_city_totals <- tree_clean %>%
  group_by(COMMUNITY) %>%
  mutate(total_trees_city = n()) %>%
  ungroup()
tree_city_totals
```


```{r, include=FALSE}
library(janitor)
tree_summary <- tree_clean %>%
  group_by(COMMUNITY) %>%
  summarise(trees = n(), .groups = "drop")
tree_summary

tree_clean %>% tabyl(COMMUNITY)
```


```{r,include=FALSE }

tree_summary_2 <- tree_summary %>%
  rename(
  City_Name = COMMUNITY
) %>%
  select(City_Name, trees)
tree_summary_2

```

```{r, include=FALSE}
# California Lakes
library(readr)
lakes <- read.csv("California_Lakes.csv")
```


```{r, include=FALSE}
# added city column
# remove NA values (no city associated with it)
# some lakes aren't associated with specific city lines so were marked as NA
library(sf)
library(tigris)
library(dplyr)

lakes_clean <- lakes %>%
  filter(!is.na(LAT_NAD83) & !is.na(LON_NAD83))

# converting to spatial points
lakes_sf <- st_as_sf(lakes_clean,
                     coords = c("LON_NAD83", "LAT_NAD83"),
                     crs = 4326)
ca_counties <- counties(state = "CA", cb = TRUE)
la_county <- ca_counties[ca_counties$NAME == "Los Angeles", ]
la_county <- st_transform(la_county, crs = st_crs(lakes_sf)) # both datasets using same coordinates before filtering

# only lakes in LA county
lakes_la <- st_filter(lakes_sf, la_county)
la_cities <- places(state = "CA", cb = TRUE)
la_cities <- st_transform(la_cities, st_crs(lakes_sf))
la_cities_la <- st_filter(la_cities, la_county)
lakes_with_city <- st_join(lakes_la, la_cities_la["NAME"])
lakes_with_city_df <- st_drop_geometry(lakes_with_city)
lakes_with_city_df
```

```{r, include=FALSE}
lakes_city_clean <- lakes_with_city %>%
  st_drop_geometry() %>%
  filter(!is.na(NAME.y))
lakes_city_clean 
```

```{r, include=FALSE}
library(stringr)
# make sure to normalize city names for join
water_cities <- lakes_city_clean %>% 
  distinct(NAME.y) %>%
  rename(City_Name = NAME.y) %>%
  mutate(
    City_Name = toupper(str_squish(City_Name)),
    lakeTF = 1
  )

water_cities
```


```{r, include=FALSE}
# Greenspace data
library(readxl)
library(dplyr)

greenspace <- read_excel("Countywide_Parks_and_Open_Space.xlsx", sheet=1)
greenspace <- greenspace %>%
  rename(City_Name = CITY) %>%
  filter(!is.na(City_Name)) %>%
  group_by(City_Name) %>%
  summarise(
    parks_in_city = n(),
    avg_acres_in_city = mean(GIS_ACRES, na.rm =TRUE),
    .groups="drop"
  ) %>%
  mutate(City_Name = toupper(str_squish(City_Name))) 
greenspace
```



```{r, include=FALSE}
# remove NAs
library(stringr)
library(dplyr)

ocean_cities <- c(
  "AVALON",
  "EL SEGUNDO",
  "HERMOSA BEACH",
  "LONG BEACH",
  "MALIBU",
  "MANHATTAN BEACH",
  "PALOS VERDES ESTATES",
  "RANCHO PALOS VERDES",
  "REDONDO BEACH",
  "SANTA MONICA"
)

LACityTractdf_3 <- LACityTractdf %>%
  mutate(
    City_Name = toupper(str_squish(City_Name))
    ) %>% 
  filter(!is.na(City_Name)) %>% 
  left_join(tree_summary_2, by = "City_Name") %>%
  left_join(water_cities, by = "City_Name") %>%
  left_join(greenspace, by = "City_Name") %>%
  mutate(
    lakeTF = ifelse(is.na(lakeTF), 0, lakeTF),
    Ocean_Feature = ifelse(
      City_Name %in% ocean_cities,1, 0)
    )
LACityTractdf_3
LACityTractdf_3 %>% tabyl(City_Name)
table(LACityTractdf_3$Ocean_Feature)
```

# Random Forest
```{r, include=FALSE}
energy_data <- LACityTractdf_3 %>%
  filter(
    HINCP.UNITS > 0) %>%
  mutate(
    total_energy_cost = ELEP.UNITS + GASP.UNITS + FULP.UNITS,
    energy_burden = total_energy_cost / HINCP.UNITS,
    energy_burden_percent = energy_burden * 100,
    
    energy_burden_cat = factor(
      ifelse(
        energy_burden < 0.06, "Low",
        ifelse(
          energy_burden <= 0.10, "High","Severe")
      ),
      levels = c("Low", "High", "Severe")
    )
  ) %>%
  filter(
    energy_burden >= 0,
    energy_burden < 1
    )

energy_data
  
summary(energy_data$total_energy_cost)
summary(energy_data$energy_burden_percent)
table(energy_data$energy_burden_cat)
```

```{r}
library(randomForest)
set.seed(123)
energy_rf <- energy_data %>%
  select(energy_burden_cat, trees, lakeTF, parks_in_city, avg_acres_in_city) %>% na.omit() %>%
  group_by(energy_burden_cat) %>% # stratified sampling accounts for uneven distribution of cats
  sample_n(20000) %>%
  ungroup()
energy_model <- randomForest(
  energy_burden_cat ~ trees + lakeTF + parks_in_city + avg_acres_in_city,
                             data = energy_rf,
                             importance = TRUE,
                             ntree = 300)
print(energy_model)
importance(energy_model)

```

Severe energy burden is predicted the best out of the three categories (based on lowest class error)
HINCP.UNITS (average annual household income) is the strongest predictor of energy burden classification (91.96). Water body presence is the least important predictor (13.17). Energy burden is heavily associated with economic factors, while environmental features contributing less.

# Descriptives of Variables

```{r}
library(skimr)
library(dplyr)

energy_data %>%
  select(energy_burden_cat, trees, lakeTF, parks_in_city, avg_acres_in_city) %>%
  skim()

energy_data %>%
  group_by(City_Name) %>%
  summarise(
    avg_trees = mean(trees, na.rm = TRUE),
    sd_trees = sd(trees, na.rm = TRUE),
    min_trees = min(trees, na.rm = TRUE),
    max_trees = max(trees, na.rm = TRUE),
    
    avg_water = mean(lakeTF, na.rm = TRUE),
    sd_water = sd(lakeTF, na.rm = TRUE),
    min_water = min(lakeTF, na.rm = TRUE),
    max_water = max(lakeTF, na.rm = TRUE),
    
    avg_parks = mean(parks_in_city, na.rm = TRUE),
    sd_parks = sd(parks_in_city, na.rm = TRUE),
    min_parks = min(parks_in_city, na.rm = TRUE),
    max_parks = max(parks_in_city, na.rm = TRUE),
    
    avg_acres = mean(avg_acres_in_city, na.rm = TRUE),
    sd_acres = sd(avg_acres_in_city, na.rm = TRUE),
    min_acres = min(avg_acres_in_city, na.rm = TRUE),
    max_acres = max(avg_acres_in_city, na.rm = TRUE)
    
  )
```

# Node Visual

```{r}
library(rpart)
library(rpart.plot)
set.seed(123)

tree_visual <- rpart(
  energy_burden_cat ~ trees + lakeTF + parks_in_city + avg_acres_in_city,
  data = energy_rf,
  method = "class"
)
rpart.plot(tree_visual)

```





```{r, include=FALSE}
# Greenspace data
library(readxl)
library(dplyr)

greenspace <- read_excel("Countywide_Parks_and_Open_Space.xlsx", sheet=1)
greenspace <- greenspace %>%
  rename(City_Name = CITY) %>%
  filter(!is.na(City_Name)) %>%
  group_by(City_Name) %>%
  summarise(
    parks_in_city = n(),
    avg_acres_in_city = mean(GIS_ACRES, na.rm =TRUE),
    .groups="drop"
  ) %>%
  mutate(City_Name = toupper(str_squish(City_Name))) 
greenspace
```






















